Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2022-03-15 ms.tex
Initial changes to address comments in the abstract and introduction

@@

\begin{abstract}
The radiation transfer equation for resonant scattering is solved in spherical geometry. A monochromatic source of photons is assumed at the center of a uniform sphere of constant gas density. The intensity within the sphere and at its surface is found in the limit of large line-center optical depth using the Eddington approximation for the angular dependence. Two limits are studied: a steady source and an impulsive source. For the steady-state problem, the solution can be represented as a sum of two terms: a previously-known analytic solution of the equation with  mean intensity $J=0$ at the surface, and a novel, semi-analytic solution which enforces the correct boundary condition of zero-ingoing intensity at the surface. This solution is compared to that of the Monte Carlo method, which is valid at arbitrary optical depth. It is shown that the size of the correction is order unity when the spectral peaks approach the Doppler core and decreases slowly with line center optical depth, which may explain discrepancies in previous studies. For the impulsive problem, the time, spatial, and frequency dependence of the solution are expressed using an eigenfunction expansion in order to characterize the escape time distribution and emergent spectra of photons. It is shown that the lowest-order eigenfrequency agrees well with the decay rate found in the Monte Carlo escape time distribution at sufficiently large line-center optical depths.

@@

The outer layers of the atmosphere are central to a planet's evolution, since they can shelter the lower atmosphere from high energy radiation as well as regulate the escape of gas into space. Given the abundance of hydrogen and its large cross section, the Lyman $\alpha$ (Ly$\alpha$) line is an important component of the radiation field. There are two sources of Ly$\alpha$: the star, and recombinations in the planet's atmosphere. \lya may ionize atoms and dissociate molecules, as well as exert pressure forces that drive an outflow. \lya can also excite H atoms to the 2p state, creating a population of Balmer-line absorbers that can be observed via transmission spectroscopy \citep{2017ApJ...851..150H}. Due to the low gas densities in the upper atmosphere, collisional de-excitation and broadening are of secondary importance and \lya may undergo ``resonant scattering''.

>>>>

\newcommand{\todo}{\textbf{TODO: }}

@@

\begin{abstract}
%ORIGINAL: 
%The radiation transfer equation for resonant scattering is solved in spherical geometry. A monochromatic source of photons is assumed at the center of a uniform sphere of constant gas density. The intensity within the sphere and at its surface is found in the limit of large line-center optical depth using the Eddington approximation for the angular dependence. Two limits are studied: a steady source and an impulsive source. For the steady-state problem, the solution can be represented as a sum of two terms: a previously-known analytic solution of the equation with  mean intensity $J=0$ at the surface, and a novel, semi-analytic solution which enforces the correct boundary condition of zero-ingoing intensity at the surface. This solution is compared to that of the Monte Carlo method, which is valid at arbitrary optical depth. It is shown that the size of the correction is order unity when the spectral peaks approach the Doppler core and decreases slowly with line center optical depth, which may explain discrepancies in previous studies. For the impulsive problem, the time, spatial, and frequency dependence of the solution are expressed using an eigenfunction expansion in order to characterize the escape time distribution and emergent spectra of photons. It is shown that the lowest-order eigenfrequency agrees well with the decay rate found in the Monte Carlo escape time distribution at sufficiently large line-center optical depths.

\textbf{We present two novel additions to the semi-analytic solution of Lyman $\alpha$ (Ly$\alpha$) radiative transfer in spherical geometry: (1) a correction to the boundary conditions for a steady source, and (2) an eigenfunction expansion to handle the time dependence of an impulsive source.} For the steady-state problem, the solution can be represented as a sum of two terms: a previously-known analytic solution of the equation with  mean intensity $J=0$ at the surface, and a novel, semi-analytic solution which enforces the correct boundary condition of zero-ingoing intensity at the surface. This solution is compared to that of the Monte Carlo method, which is valid at arbitrary optical depth. It is shown that the size of the correction is of order unity when the spectral peaks approach the Doppler core and decreases slowly with line center optical depth, \textbf{specifically $(a \tau_0)^{-1/3}$}, which may explain discrepancies in previous studies. For the impulsive problem, the time, spatial, and frequency dependence of the solution are expressed using an eigenfunction expansion in order to characterize the escape time distribution and emergent spectra of photons. It is shown that the lowest-order eigenfrequency agrees well with the decay rate found in the Monte Carlo escape time distribution at sufficiently large line-center optical depths. \textbf{The new additions presented in this work have important applications in the study of planetary atmospheres irradiated by \lya, as the correction term is significant for line-center optical depths as low as $\tau_0 \approx 10^5$ (\todo As low as? As high as? How do we stress the importance of the bc correction proprely?). Additionally, the characterization of the escape-time distribution highlights the potential for a Monte Carlo acceleration method, which would sample photon escape properties from distributions rather than calculate every photon scattering, which may be computationally expensive at large line-center optical depths.}

@@

Given the abundance of hydrogen in our universe and its large cross section, \textbf{the Lyman $\alpha$ (Ly$\alpha$) line is an important component of radiation fields in wide range of astrophysical settings. \lya radiation transport is an active area of research in the study of planets, stars, galaxies, and cosmology. In this work, however, we consider the role \lya radiation transport plays in planetary atmospheres.} The outer layers of the atmosphere are central to a planet's evolution, since they can shelter the lower atmosphere from high energy radiation as well as regulate the escape of gas into space.  There are two sources of Ly$\alpha$: the star, and recombinations in the planet's atmosphere. \lya may ionize atoms and dissociate molecules, as well as exert pressure forces that drive an outflow (\todo Citation needed). \lya can also excite H atoms to the 2p state, creating a population of Balmer-line absorbers that can be observed via transmission spectroscopy \citep{2017ApJ...851..150H}. Due to the low gas densities in the upper atmosphere, collisional de-excitation and broadening are of secondary importance and \lya may undergo ``resonant scattering''.
